Supplement to Initial estuarine response to the nutrient-rich Piney Point release into Tampa Bay, Florida

### 2021-12-06 manu-draft.Rmd
nw second revision done

Statistical assessments were evaluated only on total nitrogen (TN), chlorophyll-a (chl-a), and Secchi disk depth as a general analysis of potential patterns in eutrophication in nitrogen-limited systems. Observations for each data type were typically aggregated to the weekly or monthly scale given that sampling occurred on different days during the six month period.  Spatial comparisons were based primarily on the three areas identified in Figure \@ref(fig:map)a. Variables with log-normal distributions were log$_10$-transformed (i.e., nutrients, chl-a) prior to analysis.  For statistical tests using water quality data, only the monitoring results from FDEP were used for analysis given the consistency of sample location and collection dates. For censored data (i.e., above or below detection), Secchi observations that were visually identified on the bottom (`r sumb` of `r totb` observations in the FDEP data) were removed from analysis, whereas all other non-detects that were below laboratory standards of detection were evaluated with methods described below.

@@

Samples collected in the bay between April through September 2021 indicated that water quality conditions were unusual and outside of normal values expected for each month. A total of `r nwq` samples were collected and analyzed for chl-a, dissolved oxygen, TN, total phosphorus, total ammonia nitrogen, nitrate/nitrite, pH, salinity, Secchi depth, and temperature (Table \@ref(tab:wqsumtab)). The percentage of observations outside of the normal range (mean +/- 1 standard deviation from long-term data) varied by location and parameter.  For chl-a, `r chlabv[['1']]`\% of the observations from April through September were above the normal range for Area 1 located closest to the discharge point, whereas only `r chlabv[[2]]`\% and `r chlabv[[3]]`\% were above for Areas 2 (to the north) and 3 (to the south), respectively. Total nitrogen concentrations were above the normal range for `r tnabv[['1']]`\% of observations in Area 1, whereas concentrations were above for `r tnabv[['2']]`\% of observations in Area 2 and `r tnabv[['3']]`\% in Area 3.  Secchi observations were below the normal range for `r secblw[['1']]`\% of observations in Area 1 and for `r secblw[['2']]`\% and `r secblw[['3']]`\% of observations in Areas 2 and 3.  Notable differences were also observed for dissolved oxygen (e.g., `r dosabv[['1']]`\% were above in Area 1, `r dosabv[['2']]`\% in Area 2).  Physical parameters (salinity, temperature) and inorganic nitrogen (ammonia, nitrate/nitrite) were more often in normal range, although initial time series showed much higher concentrations for ammonia in April near Area 1.  Ammonia concentrations were observed in excess of 10 mg/L, about three orders of magnitude above baseline (see supplement, similar to the discharge measurements in Table \@ref(tab:stktab)). Inorganic nitrogen did not persist at high concentrations past April as it was likely rapidly utilized by phytoplankton (see below).  Spatial variation among the parameters showed that values were generally above the normal range (or below for Secchi depth) for many locations near Piney Point (Area 1), Anna Maria Sound (Area 3), and the northern mouth of Tampa Bay (Area 3, Figure \@ref(fig:wqmap)).

@@

Medium bloom concentrations of the red tide species *K. brevis* in 2021 were first observed in Tampa Bay the week of May 23, with high bloom concentrations peaking by the week of July 4th, after which concentrations declined (Figure \@ref(fig:redtide)b). The increase in *K. brevis* from April to July was an anomaly in 2021 that is not regularly observed in Tampa Bay.  The historical record from 1953 to present (Figure \@ref(fig:redtide)a) shows cell concentrations sampled in Tampa Bay between April and September, with only a few years having cell concentrations greater than 10,000 cells/L, notably 1963, 1971, 2005, 2018, and 2021. Median cell concentrations for most years were well below 1,000 cells/L. The highest concentration of 20 million cells/L was observed in 1971, whereas the second highest concentration of 17.6 million cells/L was observed in 2021.  Cumulative rainfall and associated inflow from the main rivers entering Tampa Bay in 2021 were below historical values (1995 - 2020) in the months preceding the highest bloom concentrations (i.e., January to June, Figure \@ref(fig:redtide)c, d). This likely contributed to elevated salinity in lower and middle Tampa Bay that created conditions favorable for *K. brevis* growth in 2021, in addition to the elevated nutrient concentrations from the Piney Point discharge.

@@

The above results can be considered together to develop a narrative of the temporal shift of nutrient pools between ecosystem components of the bay from April through September, starting with the influx of inorganic nitrogen from Piney Point. Total nitrogen concentrations first peaked in April (Figure \@ref(fig:nutrientflow)a), as did chl-a concentrations (Figure \@ref(fig:nutrientflow)b).  This initial peak in water quality parameters suggested a unique response of the phytoplankton community as an increase in diatoms (e.g., centric species, also *Asterionellopsis* sp. and *Skeletonema* sp., Figure \@ref(fig:nutrientflow)c) that can readily utilize inorganic forms of nitrogen that were a majority percentage of the TN in the initial discharge.  Water quality indicators improved slightly following the decrease in diatoms in late April, as noted by relatively lower concentrations of TN and chl-a.  However, macroalgal biomass increased after the initial diatom bloom and peaked in June (Figure \@ref(fig:nutrientflow)d), suggesting a shift of nutrients from phytoplankton to drift macroalgae communities. During peak macroalgae growth, TN and chl-a concentrations remained relatively low as nutrients were likely retained in macroalgae, until late June and early July when *K. brevis* concentrations peaked (Figure \@ref(fig:nutrientflow)e).  The co-occurring decline in macroalgae and increase in *K. brevis* suggests a release of nutrients from the former that stimulated growth of the latter, although residual water column nutrients from Piney Point were likely present (as noted by modelling efforts).  Finally, conditions were relatively stable in August and September with relatively improved water quality conditions and no dominant algal blooms.  These distinct temporal periods were readily identified through an ordination plot (see supplement) for the observed data in Figure \@ref(fig:nutrientflow).

@@

In addition to the changing environmental conditions observed during the summer, the Piney Point event also represented an anomalous volume and load of labile nitrogen released directly into lower Tampa Bay.  Spill events [reported to FDEP](https://prodenv.dep.state.fl.us/DepPNP/reports/viewIncidentDetails) (e.g., industrial spills, service line failures, sanitary sewer overflows) provide additional context for Piney Point relative to other potential anomalous releases to Tampa Bay. An assessment of over 800 reports to FDEP for the Tampa Bay watershed over the last five years showed spill volumes for these events are small (median volume 13.7 thousand liters [unpublished analysis](https://tbep-tech.github.io/piney-point-analysis/spills)) compared to the 814 million liters released from Piney Point. Moreover, the estimated nutrient load of 186 metric tons of nitrogen to Tampa Bay from Piney Point over the ten day period, exceeded current annual estimates of all external loading sources into lower Tampa Bay [@tbep0417].  External nitrogen loads to lower Tampa Bay averaged `r tnld` metric tons per year for the baseline period of 2006 to 2020 (https://tbep-tech.github.io/load-estimates/).  From both a monitoring perspective and the nature of the unanticipated release into Tampa Bay, indicators of eutrophication from the 2021 water quality data were comparatively different than recent prior years. These events were interpreted in the context of broad goals of the Tampa Bay Estuary Program and its partners to collectively manage ecosystem health following a nutrient management paradigm with seagrasses as both an indicator of environmental quality and an asset for biological resources in Tampa Bay.

@@

For seagrasses, major bloom events in 2021 produced unfavourable water quality conditions, although changes in frequency occurrence of seagrasses were minimal over the initial study period. The long-term effects of the Piney Point discharge on the seagrass community remains uncertain. From 2018 to 2020, seagrass coverage declined by 16\% in Tampa Bay, with similar losses observed in Sarasota Bay (18\%), Lemon Bay (12\%), and Charlotte Harbor (23\%) to the south (Southwest Florida Water Management District, unpublished results).  These broader trends suggest regional drivers are affecting seagrass communities [e.g., variation in precipitation, @Tomasko20], yet local issues specific to individual bays also pose challenges to managing water quality and subtidal habitats. Recent seagrass losses in Sarasota Bay may be linked to decreased light availability from a persistent *K. brevis* bloom in 2018. Although the 2021 red tide in Tampa Bay was short-lived, potential long-term effects on seagrasses remain a concern. Ecosystem shifts from seagrass to macroalgae dominated communities are also a concern, both in 2021 and as observed at some locations in recent years from the annual transect monitoring results for Tampa Bay.  In particular, increasing abundance in recent years of the green algae *Caulerpa* sp. has been observed at long-term transects that were previously dominated by seagrass.  These changes may be indicative of broader ecosystem shifts concurrent with alteration of nutrient loads or system resilience at the expense of seagrass communities [@Lloret05;@Stafford06]. Acute stressors from short-term events, such as unanticipated releases from Piney Point, create additional and often preventable challenges in the context of broader issues affecting ecosystem health.

@@

In the broader context of mining impacts to surface waters, the results reinforce the understanding that legacy pollutants from phosphate mining can negatively affect environmental resources.  In addition to the previous studies and reports for Tampa Bay [@Garrett11;@Switzer11], other Gulf Coast estuaries have been affected by pollutants from unanticipated gypstack releases.  Grand Bay is a 7500 hectare protected area in southern Mississippi that has been exposed to phosphorus-rich and highly acidic water from a defunct gypstack [@Dillon15;@Beck18b].  Two spills have occurred in Grand Bay, the first in 2005 following failure of the retaining walls after a heavy rain event and the second in 2012 after passage of Hurricane Isaac when the holding capacity of the gypstack was exceeded again with heavy rainfall.  Massive fish kills were observed and likely related to low pH of the water released.  Unlike Piney Point, inorganic nitrogen concentrations of the release were low due to a different fertilizer production method and concerns of the long-term impacts focused primarily on heavy loads of orthophosphate [@Dillon15].  Phosphate loads to Tampa Bay from Piney Point were similar in magnitude to the nitrogen loads, although concentrations were within normal baseline ranges within a month after the release stopped (see supplement).  The fate of the phosphorus in Tampa Bay is unknown.  Regardless, the historical context of Grand Bay is similar to Piney Point and other international examples, e.g., Huelva estuary in Spain [@Perez10;@Perez16].  Legacy wastewater from fertilizer production has been poorly maintained at some facilities and long-term plans are insufficient to safely dispose of remnant pollutants that pose a risk of significant impacts to coastal resources that increases over time.  These are not isolated examples and enhanced regulatory oversight is needed to safely and effectively close these types of facilities [@Nelson21].

@@

```{r stktab}
cap.val <- 'Measured concentrations from the phosphogypsum stack (NGS-S) at Piney Point from a 2019 sample and end-of-pipe samples from April 2021 for relevant water quality variables.  Values are compared to normal annual medians (min, max) for concentrations in lower Tampa Bay.  Normal medians are based on data for a baseline period from 2006 to 2020 from long-term monitoring stations in lower Tampa Bay (Figure \\@ref(fig:map)a).  End-of-pipe concentrations were taken as averages from two samples on April 6th and 13th at the outflow at Port Manatee.  Missing values were not measured in the stack water or release water.'
knitr::kable(stktab, booktabs = T, caption = cap.val)
```

>>>>

Statistical assessments were evaluated only on total nitrogen (TN), chlorophyll-a (chl-a), and Secchi disk depth as a general analysis of potential patterns in eutrophication in nitrogen-limited systems. Observations for each data type were typically aggregated to the weekly or monthly scale given that sampling occurred on different days during the six month period.  Spatial comparisons were based primarily on the three areas identified in Figure \@ref(fig:map)a. Variables with log-normal distributions were log$_10$-transformed (i.e., nutrients, chl-a) prior to analysis.  For statistical tests using water quality data, only the monitoring results from FDEP were used for analysis given the consistency of sample location and collection dates. Secchi observations that were visually identified on the bottom (`r sumb` of `r totb` observations in the FDEP data) were removed from analysis. Observations for other parameters that were below laboratory standards of detection were evaluated with methods described below.

@@

Samples collected in the bay between April through September 2021 indicated that water quality conditions were unusual and outside of normal values expected for each month. A total of `r nwq` samples were collected and analyzed for chl-a, dissolved oxygen, TN, total phosphorus, total ammonia nitrogen, nitrate/nitrite, pH, salinity, Secchi depth, and temperature (Table \@ref(tab:wqsumtab)). The percentage of observations outside of the normal range (mean +/- 1 standard deviation from long-term data) varied by location and parameter.  For chl-a, `r chlabv[['1']]`\% of the observations from April through September were above the normal range for Area 1 located closest to the discharge point, whereas only `r chlabv[[2]]`\% and `r chlabv[[3]]`\% were above for Areas 2 (to the north) and 3 (to the south), respectively. Total nitrogen concentrations were above the normal range for `r tnabv[['1']]`\% of observations in Area 1, whereas concentrations were above for `r tnabv[['2']]`\% of observations in Area 2 and `r tnabv[['3']]`\% in Area 3.  Secchi observations were below the normal range for `r secblw[['1']]`\% of observations in Area 1 and for `r secblw[['2']]`\% and `r secblw[['3']]`\% of observations in Areas 2 and 3.  Notable differences were also observed for dissolved oxygen (e.g., `r dosabv[['1']]`\% were above in Area 1, `r dosabv[['2']]`\% in Area 2).  Physical parameters (salinity, temperature) and inorganic nitrogen (ammonia, nitrate/nitrite) were more often in normal range, although initial time series showed much higher concentrations for ammonia in April near Area 1.  Ammonia concentrations near the point of discharge were observed in excess of 10 mg/L, about three orders of magnitude above baseline (see supplement, similar to the discharge measurements in Table \@ref(tab:stktab)). Inorganic nitrogen did not persist at high concentrations past April as it was likely rapidly utilized by phytoplankton (see below).  Spatial variation among the parameters showed that values were generally above the normal range (or below for Secchi depth) for many locations near Piney Point (Area 1), Anna Maria Sound (Area 3), and the northern mouth of Tampa Bay (Area 3, Figure \@ref(fig:wqmap)).

@@

Medium bloom concentrations of the red tide species *K. brevis* in 2021 were first observed in Tampa Bay the week of May 23, with high bloom concentrations peaking by the week of July 4th, after which concentrations declined (Figure \@ref(fig:redtide)b). The increase in *K. brevis* from April to July was an anomaly in 2021 that is not regularly observed in Tampa Bay.  The historical record from 1953 to present (Figure \@ref(fig:redtide)a) shows cell concentrations sampled in Tampa Bay between April and September, with only a few years having cell concentrations greater than 10,000 cells/L, notably 1963, 1971, 2005, 2018, and 2021. Median cell concentrations for most years were well below 1,000 cells/L. The highest concentration of 20 million cells/L was observed in 1971, whereas the second highest concentration of 17.6 million cells/L was observed in 2021.  Cumulative rainfall and associated inflow from the main rivers entering Tampa Bay in 2021 were below historical values (1995 - 2020) in the months preceding the highest bloom concentrations (i.e., January to June, Figure \@ref(fig:redtide)c, d). This likely contributed to elevated salinity in lower and middle Tampa Bay that created conditions favorable for *K. brevis* growth in 2021 (Figure S2f, S3f), in addition to the elevated nutrient concentrations from the Piney Point discharge.

@@

The above results can be considered together to develop a narrative of the temporal shift of nutrient pools between ecosystem components of the bay from April through September, starting with the influx of inorganic nitrogen from Piney Point. Total nitrogen concentrations first peaked in April (Figure \@ref(fig:nutrientflow)a), as did chl-a concentrations (Figure \@ref(fig:nutrientflow)b).  This initial peak in water quality parameters suggested a unique response of the phytoplankton community as an increase in diatoms (e.g., centric species, also *Asterionellopsis* sp. and *Skeletonema* sp., Figure \@ref(fig:nutrientflow)c) that can readily utilize inorganic forms of nitrogen that were present in the initial discharge [@Bates76;@Domingues11].  Water quality indicators improved slightly following the decrease in diatoms in late April, as noted by relatively lower concentrations of TN and chl-a.  However, macroalgal biomass increased after the initial diatom bloom and peaked in June (Figure \@ref(fig:nutrientflow)d), suggesting a shift of nutrients from phytoplankton to drift macroalgae communities. During peak macroalgae growth, TN and chl-a concentrations remained relatively low as nutrients were likely retained in macroalgae, until late June and early July when *K. brevis* concentrations peaked (Figure \@ref(fig:nutrientflow)e).  The co-occurring decline in macroalgae and increase in *K. brevis* suggests a release of nutrients from the former that stimulated growth of the latter, although residual water column nutrients from Piney Point may have also been present (as suggested by modelling efforts).  Finally, conditions were relatively stable in August and September with relatively improved water quality conditions and no dominant algal blooms.  These distinct temporal periods were readily identified through an ordination plot (see supplement) for the observed data in Figure \@ref(fig:nutrientflow).

@@

In addition to the changing environmental conditions observed during the summer, the Piney Point event also represented an anomalous volume and load of labile nitrogen released directly into lower Tampa Bay.  Spill events [reported to FDEP](https://prodenv.dep.state.fl.us/DepPNP/reports/viewIncidentDetails) (e.g., industrial spills, service line failures, sanitary sewer overflows) provide additional context for Piney Point relative to other potential anomalous releases to Tampa Bay. An assessment of over 800 reports to FDEP for the Tampa Bay watershed over the last five years showed spill volumes for these events are small (median volume 13.7 thousand liters [unpublished analysis](https://tbep-tech.github.io/piney-point-analysis/spills)) compared to the 814 million liters released from Piney Point. Moreover, the estimated nutrient load of 186 metric tons of nitrogen to Tampa Bay from Piney Point over the ten day period, exceeded current annual estimates of all external loading sources into lower Tampa Bay [@tbep0417].  External nitrogen loads to lower Tampa Bay averaged `r tnld` metric tons per year for the baseline period of 2006 to 2020 (https://tbep-tech.github.io/load-estimates/).  As anticipated, the 2021 results showed that indicators of eutrophication from the water quality data were comparatively different than recent prior years. These events were interpreted in the context of broad goals of the Tampa Bay Estuary Program and its partners to collectively manage ecosystem health following a nutrient management paradigm with seagrasses as both an indicator of environmental quality and an asset for biological resources in Tampa Bay.

@@

For seagrasses, major bloom events in 2021 produced unfavourable water quality conditions, although changes in frequency occurrence of seagrasses were not observed over the initial study period. The long-term effects of the Piney Point discharge on the seagrass community remains uncertain. From 2018 to 2020, seagrass coverage declined by 16\% in Tampa Bay, with similar losses observed in Sarasota Bay (18\%), Lemon Bay (12\%), and Charlotte Harbor (23\%) to the south (Southwest Florida Water Management District, unpublished results).  These broader trends suggest regional drivers are affecting seagrass communities [e.g., variation in precipitation, @Tomasko20], yet local issues specific to individual bays also pose challenges to managing water quality and subtidal habitats. Recent seagrass losses in Sarasota Bay may be linked to decreased light availability from a persistent *K. brevis* bloom in 2018. Although the 2021 red tide in Tampa Bay was short-lived, potential long-term effects on seagrasses remain a concern. Ecosystem shifts from seagrass to macroalgae dominated communities are also a concern, both in 2021 and as observed at some locations in recent years from the annual transect monitoring results for Tampa Bay.  In particular, increasing abundance in recent years of the green algae *Caulerpa* sp. has been observed at long-term transects that were previously dominated by seagrass.  These changes may be indicative of broader ecosystem shifts concurrent with alteration of nutrient loads or system resilience at the expense of seagrass communities [@Lloret05;@Stafford06]. Acute stressors from short-term events, such as unanticipated releases from Piney Point, create additional and often preventable challenges in the context of broader issues affecting ecosystem health.

@@

In the broader context of mining impacts to surface waters, the results reinforce the understanding that legacy pollutants from phosphate mining can negatively affect environmental resources.  In addition to the previous studies and reports for Tampa Bay [@Garrett11;@Switzer11], other Gulf Coast estuaries have been affected by pollutants from unanticipated gypstack releases.  Grand Bay is a 7500 hectare protected area in southern Mississippi that has been exposed to phosphorus-rich and highly acidic water from a defunct gypstack [@Dillon15;@Beck18b].  Two spills have occurred in Grand Bay, the first in 2005 following failure of the retaining walls after a heavy rain event and the second in 2012 after passage of Hurricane Isaac when the holding capacity of the gypstack was exceeded again with heavy rainfall.  Massive fish kills were observed and likely related to low pH of the water released.  Unlike Piney Point, inorganic nitrogen concentrations of the release were low due to a different fertilizer production method and concerns of the long-term impacts focused primarily on heavy loads of orthophosphate [@Dillon15].  Phosphate loads to Tampa Bay from Piney Point were similar in magnitude to the nitrogen loads, although concentrations were within normal baseline ranges within a month after the release stopped (see supplement).  The fate of the phosphorus in Tampa Bay is less understood than that of nitrogen.  Regardless, the historical context of Grand Bay is similar to Piney Point and other international examples, e.g., Huelva estuary in Spain [@Perez10;@Perez16].  Legacy wastewater from fertilizer production has been poorly maintained at some facilities and long-term plans are insufficient to safely dispose of remnant pollutants that pose a risk of significant impacts to coastal resources that increases over time.  These are not isolated examples and enhanced regulatory oversight is needed to safely and effectively close these types of facilities [@Nelson21].

@@

```{r stktab}
cap.val <- 'Measured concentrations from the phosphogypsum stack (NGS-S) at Piney Point from a 2019 sample and samples from April 2021 for relevant water quality variables.  Values are compared to normal annual medians (min, max) for concentrations in lower Tampa Bay.  Normal medians are based on data for a baseline period from 2006 to 2020 from long-term monitoring stations in lower Tampa Bay (Figure \\@ref(fig:map)a).  The 2021 samples are from the NGS-S stack on April 13th and directly from the outflow site at Port Manatee on April 6th. Missing values were not measured in the stack water or release water.'
knitr::kable(stktab, booktabs = T, caption = cap.val)
```

### 2021-12-06 supplement.Rmd
nw second revision done

Established laboratory and field sample protocols for all survey methods were based on an [Interagency Monitoring Project Plan](https://drive.google.com/drive/u/0/folders/1oBGvjdve-Gpo4Kn3Ovn8a8-yVoP25eec) maintained by the TBEP in agreement with USEPA standards and those of the inter-agency partners.  Data quality objectives followed guidelines outlined in the TBEP Data Quality Management Plan [@tbep1620]. Many of the local partners also participate in the Southwest Florida [Regional Ambient Monitoring Program](https://tbep.org/our-work/boards-committees/technical-advisory-committee/#ramp) (RAMP) that ensures similar standards and protocols are followed in the collection and processing of monitoring data, including routine cross-reference of samples between laboratories to check precision of measured values.  Discrete water quality samples were taken primarily from surface grabs by boat and processed for relevant parameters (see below) by the respective laboratories of each participating agency.  Surface samples requiring laboratory analysis (e.g., nutrient assays) were obtained primarily from bottle collection at the surface, whereas in situ measurements were available for many parameters (e.g., dissolved oxygen, secchi depth, etc.).  In situ measurements were collected using common monitoring equipment, such as YSI sondes or Seabird CTD casts depending on monitoring agency.  Laboratory methods used to process samples were based on accepted procedures promoted through the Southwest Florida RAMP.

For data analysis in this paper, we focus on water quality parameters related to the Tampa Bay nutrient management paradigm and the expected phytoplankton response from a dense, inorganic nitrogen plume entering the bay.  This included statistical evaluation of total nitrogen (mg/L), total ammonia nitrogen (NH$_3$ + NH$_4^+$, mg/L), nitrate/nitrite (NO$_3^-$ + NO$_2^-$, mg/L), total phosphorus (mg/L), orthophosphate (PO$_4^{3-}$, mg/L), and chlorophyll-a (ug/L) concentrations.  Samples for pH, salinity (psu), temperature (C), and dissolved oxygen saturation (\%) were also evaluated given the role these parameters can have as indicators of surface water contamination (pH), physical drivers of primary production (salinity, temperature), and indicators of primary production and respiration (dissolved oxygen, mg/L).  Overall, sample frequency was approximately biweekly over the subsequent months, with some variability given agency resource limitations during the event and weather restrictions on boating.  As appropriate, water quality data were aggregated at the weekly scale and by major areas of interest (Figure 1a) given the hypothesized impacts adjacent to the release site as indicated by the TBCOM tracer model simulated plume patterns.

>>>>

Established laboratory and field sample protocols for all survey methods were based on an [Interagency Monitoring Project Plan](https://drive.google.com/drive/u/0/folders/1oBGvjdve-Gpo4Kn3Ovn8a8-yVoP25eec) maintained by the TBEP and those of the inter-agency partners.  Data quality objectives followed guidelines outlined in the USEPA-approved TBEP Data Quality Management Plan [@tbep1620]. Many of the local partners also participate in the Southwest Florida [Regional Ambient Monitoring Program](https://tbep.org/our-work/boards-committees/technical-advisory-committee/#ramp) (RAMP) that ensures similar standards and protocols are followed in the collection and processing of monitoring data, including routine cross-reference of samples between laboratories to check precision of measured values.  Discrete water quality samples were taken primarily from surface grabs by boat and processed for relevant parameters (see below) by the respective laboratories of each participating agency.  Surface samples requiring laboratory analysis (e.g., nutrient assays) were obtained primarily from bottle collection at the surface, whereas in situ measurements were available for many parameters (e.g., dissolved oxygen, secchi depth, etc.).  In situ measurements were collected using common monitoring equipment, such as YSI sondes or Seabird CTD casts depending on monitoring agency.  Laboratory methods used to process samples were based on accepted procedures promoted through the Southwest Florida RAMP.

For data analysis in this paper, we focus on water quality parameters related to the Tampa Bay nutrient management paradigm and the expected phytoplankton response from a dense, inorganic nitrogen plume entering the bay.  This included statistical evaluation of total nitrogen (mg/L), total ammonia nitrogen (NH$_3$ + NH$_4^+$, mg/L), nitrate/nitrite (NO$_3^-$ + NO$_2^-$, mg/L), total phosphorus (mg/L), orthophosphate (PO$_4^{3-}$, mg/L), and chlorophyll-a (ug/L) concentrations.  Samples for pH, salinity (psu), temperature (C), and dissolved oxygen saturation (\%) were also evaluated given the role these parameters can have as indicators of surface water contamination (pH), physical drivers of primary production (salinity, temperature), and indicators of primary production and respiration (dissolved oxygen).  Overall, sample frequency was approximately biweekly over the subsequent months with greatest effort during the first three months and with some variability given agency resource limitations during the event and weather restrictions on boating.  As appropriate, water quality data were aggregated at the weekly scale and by major areas of interest (Figure 1a) given the hypothesized impacts adjacent to the release site as indicated by the TBCOM tracer model simulated plume patterns.
